---
title: "**Exploring Key Drivers of House Prices Through Regression-Based Analysis**"
author: '**Group 4**'
date: "**2025-04-23**"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
## Introduction and Background



Understanding what drives *house prices* is essential for both *buyers and sellers, yet in many cases, especially in **less transparent markets, people struggle to determine a **fair price* for a property with specific features. This can make home buying or selling a *confusing and risky* process.

*Data analysis* offers a solution by revealing how *house characteristics—like size, number of rooms, or location—relate to price. This project uses **open-source housing data* and applies *multiple linear regression* to identify key factors influencing house prices.

Although the dataset is based on homes in the *United States, the findings and approach can be valuable in other contexts where **real estate pricing* remains unclear. The goal is to build a simple, *data-driven model* that can help estimate house prices based on common property features.

 <p align="center"><strong>Research Question and Goals for Data Analysis</strong></p>

This project began with an *exploratory data analysis (EDA)* to better understand the dataset and identify patterns or relationships between house features and sale prices. During the EDA, we observed that variables such as *living area, **number of bathrooms, **grade, and **renovation status* appeared to have strong associations with price. These early insights guided our decision to build a predictive model.

<p><strong>Research Question</strong></p>

*What are the key housing features that significantly influence house sale prices, and can we build a regression model to predict price based on those features?*

<p><strong>Goals of the Analysis</strong></p>

1. *Explore relationships* between house characteristics and sale price using visual and statistical techniques.  
2. *Identify significant predictors* of house price through inferential methods.  
3. *Build and interpret* a multiple linear regression model to estimate house prices based on selected features.  
4. *Provide insights* that can support decision-making for individuals looking to buy or sell a home, using data instead of guesswork.

```{r,echo=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(maps)
library(corrplot)
data <- read.csv("C:/Users/user/Documents/new_project/traffic/data.csv")
```
1) EDA

```{r,echo=FALSE}
glimpse(data)
```



```{r}
summary(data)
```
```{r}
str(data)
```

```{r}
head(data)
```


```{r}
colnames(data)
```


```{r}
sum(is.na(data))
```
```{r}
duplicated_rows <- duplicated(data)
sum(duplicated(data)) 
```
2) INSIGHTS

```{r}

# Create the box plot
boxplot(price ~ bedrooms,
        data = data,
        main = "Box Plot of Price by Number of Bedrooms",
        xlab = "Number of Bedrooms",
        ylab = "Price",
        col = "lightgreen",
        las = 1)  # Horizontal y-axis labels

```


```{r}

library(dplyr)
library(ggplot2)

data %>%
  group_by(city) %>%
  summarise(avg_price = mean(price)) %>%
  top_n(10, avg_price) %>%
  ggplot(aes(x = reorder(city, avg_price), y = avg_price)) +
  geom_col(fill = "#69b3a2") +
  coord_flip() +
  labs(title = "Top 10 Cities by Avg Price", x = "City", y = "Avg Price ($)") +
  theme_minimal()




```

```{r}
plot(data$sqft_living, data$price,
     main = "Price vs Sqft Living",
     xlab = "Sqft Living",
     ylab = "Price",
     pch = 19,
     col = "darkblue")
abline(lm(price ~ sqft_living, data = data), col = "red", lwd = 2)


```
```{r}
plot(data$sqft_lot, data$price,
     main = "Price vs Sqft Lot",
     xlab = "Sqft Lot",
     ylab = "Price",
     pch = 19,
     col = "green")
abline(lm(price ~ sqft_lot, data = data), col = "red", lwd = 2)


```
```{r}
ggplot(data,aes(y=price,x=sqft_living)) +
       geom_point() + 
        xlim(0, 9000) +
        ylim(0, 5000000) +
        geom_smooth(formula = y ~ x,method="lm")
```


```{r}
library(car)
# Remove rows with missing or problematic price values
clean_data <- data[!is.na(data$price) & data$price > 0, ]

# Fit the model with clean data
model <- lm(log(price) ~ sqft_living + bedrooms + bathrooms + sqft_lot + yr_built + condition + yr_renovated +waterfront + sqft_above,
data = clean_data)
summary(model)

```
```{r}
model <- lm(log(price) ~ log(sqft_living)+ bedrooms + bathrooms + sqft_lot +
              yr_built + condition + yr_renovated + waterfront + view +
              sqft_above + sqft_basement + floors + city,
            data = clean_data)
summary(model)


```
```{r}
stepwise_model <- step(model, direction = "both")
summary(stepwise_model)
```


```{r}
summary(model)$r.squared
```
```{r}
# Extract residuals and fitted values
 residuals <- resid(model)
 fitted_values <- fitted(model)
 # 1. Residuals vs. Fitted Plot
 plot(fitted_values, residuals,
 main = "Residuals vs Fitted Values",
 xlab = "Fitted Values",
 ylab = "Residuals",
 pch = 19, col = "blue")
 abline(h = 0, lty = 2, col = "red")


```

```{r}


numeric_data <- select_if(clean_data, is.numeric)
corr_matrix <- cor(numeric_data, use = "complete.obs")
corrplot(corr_matrix, method = "color", type = "upper", tl.cex = 0.8)



```



```{r}
ggplot(clean_data, aes(x = factor(condition), y = price)) +
  geom_boxplot(fill = "lightcoral") +
  labs(title = "Price Distribution by Condition", x = "Condition", y = "Price") +
  theme_minimal()

```
```{r}
library(ggplot2)
ggplot(clean_data, aes(x = sqft_living, y = price)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  geom_smooth(method = "lm", color = "red") +
  facet_wrap(~ condition, scales = "free_y") +
  labs(title = "Price vs. Living Area by House Condition",
       x = "Living Area (sqft)", y = "Price ($)") +
  theme_minimal()
```

```{r}
library(dplyr)
library(ggplot2)
# Summarize average price by year built
price_by_year <- clean_data %>%
  group_by(yr_built) %>%
  summarise(avg_price = mean(price, na.rm = TRUE)) %>%
  ungroup()
# Create line plot
ggplot(price_by_year, aes(x = yr_built, y = avg_price)) +
  geom_line(color = "darkblue", size = 1) +
  geom_point(color = "darkblue", size = 2) +
  labs(title = "Average House Price by Year Built",
       x = "Year Built", y = "Average Price ($)") +
  theme_minimal()
```
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)

# Load and clean the data
data <- read.csv("data.csv")

# Convert date to Date format and filter valid prices
clean_data <- data %>%
  mutate(date = as.Date(date)) %>%
  filter(!is.na(price) & price > 0)

# Aggregate prices by date (mean price per day)
price_by_date <- clean_data %>%
  group_by(date) %>%
  summarise(avg_price = mean(price, na.rm = TRUE)) %>%
  ungroup()

# Create the line chart
ggplot(price_by_date, aes(x = date, y = avg_price)) +
  geom_line(color = "darkblue", size = 1) +
  geom_point(color = "darkblue", size = 2) +
  labs(title = "Average House Prices Over Time (May-July 2014)",
       x = "Date",
       y = "Average Price ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
# Step 1: Load the data into a data frame called 'data'
data <- read.csv("data.csv", stringsAsFactors = FALSE)

# Step 2: Inspect the data
cat("First few rows of the dataset:\n")
head(data)

cat("\nStructure of the dataset:\n")
str(data)

# Step 3: Compute descriptive statistics for numeric columns
cat("\nDescriptive statistics for numeric columns:\n")
numeric_cols <- sapply(data, is.numeric)
desc_stats <- summary(data[, numeric_cols])
print(desc_stats)

# Alternative: Use psych package for more detailed statistics
# Install psych if not already installed: install.packages("psych")
library(psych)
cat("\nDetailed descriptive statistics (using psych package):\n")
psych_stats <- describe(data[, numeric_cols])
print(psych_stats)

# Step 4: Summarize categorical columns
cat("\nSummary for categorical columns:\n")
categorical_cols <- sapply(data, is.character)
for (col in names(data)[categorical_cols]) {
  cat("\nColumn:", col, "\n")
  print(table(data[[col]]))
}

# Optional: Use summarytools for a comprehensive summary
# Install summarytools if not already installed: install.packages("summarytools")
library(summarytools)
cat("\nComprehensive summary (using summarytools):\n")
print(dfSummary(data))
```

```{r}
# Debugging: Inspect the city and condition columns
cat("Unique values in city:\n")
print(unique(data$city))
cat("\nUnique values in condition:\n")
print(unique(data$condition))
cat("\nSummary of city:\n")
print(summary(as.factor(data$city)))
cat("\nSummary of condition:\n")
print(summary(as.factor(data$condition)))

# Selecting numeric columns for descriptive statistics
numeric_cols <- data %>%
  select(price, bedrooms, bathrooms, sqft_living, sqft_lot, floors,
         sqft_above, sqft_basement, yr_built, yr_renovated)

# Calculating descriptive statistics for numeric variables
numeric_stats <- descr(numeric_cols, stats = c("mean", "sd", "min", "q1", "med", "q3", "max"), round.digits = 2)

# Printing numeric statistics
cat("\nDescriptive Statistics for Numeric Variables:\n")
print(numeric_stats)
```


```{r}


# Step 2: Inspect the data
cat("First few rows of the dataset:\n")
head(data)

cat("\nStructure of the dataset:\n")
str(data)

# Step 3: Compute descriptive statistics for numeric columns
cat("\nDescriptive statistics for numeric columns:\n")
numeric_cols <- sapply(data, is.numeric)
desc_stats <- summary(data[, numeric_cols])
print(desc_stats)

# Alternative: Use psych package for more detailed statistics
# Install psych if not already installed: install.packages("psych")
library(psych)
cat("\nDetailed descriptive statistics (using psych package):\n")
psych_stats <- describe(data[, numeric_cols])
print(psych_stats)

# Step 4: Summarize categorical columns
cat("\nSummary for categorical columns:\n")
categorical_cols <- sapply(data, is.character)
for (col in names(data)[categorical_cols]) {
  cat("\nColumn:", col, "\n")
  print(table(data[[col]]))
}

# Optional: Use summarytools for a comprehensive summary
# Install summarytools if not already installed: install.packages("summarytools")
library(summarytools)
cat("\nComprehensive summary (using summarytools):\n")
print(dfSummary(data))
```

```{r}
png("correlation_matrix.png", width = 800, height = 600)
corrplot(corr_matrix, method = "color", type = "upper", tl.cex = 0.8, title = "Correlation Matrix")
dev.off()
```

